---
title: "H4H SADC Livestock Emissions Reduction Estimation"
author: "Ruan de Wet"
date: "02/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

This analysis should clip, wrangle and process the [FAO Cattle data](http://www.fao.org/livestock-systems/global-distributions/cattle/en/) required for the H4H SADC Livestock Emissions project. 

The countries of interest include all non-SID SADC countries. Or those that have an appreciable number of cattle. There may be cause to include something about the sheep as well. 

*One red flag to note: Jacques van Rooyen has yet to respond to my messages to him. Sarah is paying half the bill, so I am proceeding with the analysis.*

## Approach

I'm using the distributions that were modelled using the dasymetric method, because I'm wanting to extract the estimated cattle numbers in spatially explicit areas (defined by the communal/commercial tenure maps and the FMD zones).

The land tenure and FMD zone shapefiles are to be provided by Jacques van Rooyen.

## Analysis

### Project Setup

```{r Load libraries and helper functions, message=FALSE}

library("sp")
library("sf")
library("tmap")
library("rgeos")
library("rgdal")
library("raster")
library("rnaturalearth")

library(RColorBrewer)
library(viridisLite)

library("ggplot2")
library("tidyr")
library("dplyr")
library("stringr")

library("readxl")
library("readr")

 # rm(list = ls())

today = Sys.Date() %>%
  format("%y%m%d_")

st_plot = function(x, ...) {x %>% st_geometry() %>% plot(...)}

```

### Define the project extent

First step is to define the project extent (i.e. the countries to be included). 

I have a .csv file in my reference data folder with the ISO codes of each country ([source](https://www.iban.com/country-codes)). That will be useful for this analysis. I also have a list of SADC countries ([source](https://en.wikipedia.org/wiki/Southern_African_Development_Community)) saved in the project directory, labelled based on whether or not it is a small island developing state. 

```{r Import spatial base info}

# Paths
dsn_data = "G:/Ruan/R/_Reference Data/"

# Metadata of consequence
countries_aoi = read_csv("countries_sadc.csv") %>% filter(SID != TRUE)
iso_all = read_excel(paste0(dsn_data, "iso_countries.xlsx")) 

# Filter down to area of interest (countries to be included)
iso_aoi = iso_all %>%
  filter(Country %in% countries_aoi$Country) 

# Not all countries were identified perfectly. The DRC and Tanzania were missed. The following code picks them out and adds them to the iso_aoi dataframe.

iso_add = iso_all %>%
  filter(str_detect(Country, "Congo") | str_detect(Country, "Tanzania"),
         ISO_3 != "COG")

iso_aoi = rbind(iso_aoi, iso_add)

```

With the ISO codes, we can easily specify the countries that are included in the study. They are loaded in the next chunk from the `rnaturalearth` package. 

```{r Load the country borders}

# Define boundary box for AOI (because there are South African Islands far South)
bound_aoi = ne_countries(continent = "africa", scale = "small",
                        returnclass = "sf", type = "sovereignty") %>%
  filter(sov_a3 %in% iso_aoi$ISO_3) %>%
  st_buffer(dist = 1) %>%
  st_union() %>%
  st_make_grid(n = 1) %>% 
    st_as_sf()

# Load the country borders of interest
bord_sadc = ne_countries(continent = "africa", scale = "medium", 
                        returnclass = "sf", type = "countries") %>%
  filter(sov_a3 %in% iso_aoi$ISO_3) 
  
# Filter borders to boundary
bord_aoi = st_intersection(bound_aoi, bord_sadc) 

```

### Process the FAO livestock data

Now that we have the countries in the study, we can process the global livestock dataset. 

```{r Import livestock data}

cattle_aoi = paste0(dsn_data, "FAO Cattle Global Distribution/5_Ct_2010_Da.tif") %>% 
  raster() %>%
  raster::crop(bound_aoi)

cattle_aoi_aw = paste0(dsn_data, "FAO Cattle Global Distribution/6_Ct_2010_Aw.tif") %>% 
  raster() %>%
  raster::crop(bound_aoi)

```

In order to extract and analyse the data appropriately, we need to use a projected CRS. Becuase of the spread in latitudes (approx. 0 to -35 degrees), I am going to use the Interrupted Goode Homolosine (IGH) projection. 

```{r Reproject data}

cattle_aoi_igh = cattle_aoi %>% raster::projectRaster(crs = "+proj=igh")
bord_aoi_igh = bord_aoi %>% st_transform(crs = "+proj=igh") %>% st_as_sf()

cattle_aoi_aw_igh = cattle_aoi_aw %>% raster::projectRaster(crs = "+proj=igh")

```

```{r Save the rasters, eval = F}

writeRaster(cattle_aoi, "H4H_Cattle SADC distribution_lonlat.tif")
writeRaster(cattle_aoi_igh, "H4H_Cattle SADC distribution_igh.tif")

writeRaster(cattle_aoi_aw, "H4H_Cattle SADC distribution_Aw_lonlat.tif")
writeRaster(cattle_aoi_aw_igh, "H4H_Cattle SADC distribution_Aw_igh.tif")

```



### Explore the data

```{r}

pal_YlOrBr = c("#cccccc", brewer.pal(n = 9, name = "YlOrBr"))

tm_shape(cattle_aoi) +
  tm_grid() +
  # tm_raster(palette = pal_YlOrBr,
  tm_raster(palette = "YlOrBr",
            style = "fixed", title = "Cattle population / km",
            breaks = c(0, 10, 20, 50, 100, 250, 500,  1000, 1500, 5000, 2*10^4, 10^5)) +
tm_shape(bord_aoi) +
  tm_borders(col = "black") +
tm_layout(legend.outside = T, bg.color = "#333333")

```

```{r}

tm_shape(cattle_aoi_aw) +
  tm_grid() +
  # tm_raster(palette = pal_YlOrBr,
  tm_raster(palette = "YlOrBr",
            style = "fixed", title = "Cattle population / km",
            breaks = c(0, 10, 20, 50, 100, 250, 500,  1000, 1500, 5000, 2*10^4, 10^5)) +
tm_shape(bord_aoi) +
  tm_borders(col = "black") +
tm_layout(legend.outside = T, bg.color = "#333333")
  
```